Bayesian Rate Estimation

### 2012-09-04 many.tex
More minor edits to 2.

For the usual $k=1$ case \citep{Biswas2009}, the marginalised
posterior (Eq.~\eqref{eq:rate-shape-posterior}) becomes
\begin{equation}
p_{\rm LE}\left( R_f, R_b, \theta | d \right) \propto \left(R_f
\hat{f}(x_N,\theta) + R_b \hat{b}(x_N,\theta)\right)\exp\left[-\left(
  R_f (1-\hat{F}(x_N,\theta)) + R_b (1-\hat{B}(x_N,\theta)) \right)
  \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
where $x_N$ denotes the loudness of the loudest event. Marginalising
over $R_b$ we obtain
\begin{equation}
p_{\rm LE}\left( R_f, \theta | d \right) \propto \left(
\frac{\hat{b}(x_N,\theta)}{2(1-\hat{B}(x_N,\theta))} + R_f
\hat{f}(x_N,\theta)\right)\frac{\sqrt{\pi}\,
  p(\theta)}{\sqrt{1-\hat{B}(x_N,\theta)}\,\sqrt{R_f}}
\exp\left(-R_f(1-\hat{F}(x_N,\theta))\right) .
\end{equation}
This posterior has a maximum in $R_f$ at
\begin{equation}
R_f =
\frac{\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta)\tilde{b}(x_N,\theta)
  \pm
  \sqrt{(\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta)\tilde{b}(x_N,\theta))^2
    -
    4\tilde{b}(x_N,\theta)(1-\hat{F}(x_N,\theta))\hat{f}(x_N,\theta)}}{4
  \hat{f}(x_N,\theta)(1-\hat{F}(x_N,\theta)}.
\end{equation}
where $\tilde{b}(x_N,\theta)
=\hat{b}(x_N,\theta)/(1-\hat{B}(x_N,\theta))$. If
$\tilde{b}(x_N,\theta) \ll 1$, we obtain the result $(1-
\hat{F}(x_N,\theta))\,R_f \approx 1/2$. This is can be understood as
the statement that the rate of foreground events with ranking
statistic greater than $x_N$, $(1- \hat{F}(x_N,\theta))\,R_f$, is of
order $1$, as expected. However, $\tilde{b}(x_N,\theta) = -{\rm d}
[\ln(1-\hat{B}(x,\theta))]/{\rm d}x$ and $(1-\hat{B}(x,\theta))
\rightarrow 0$ as $x \rightarrow \infty$, so this term will be
divergent and for many reasonable examples, we will find
$\tilde{b}(x_N,\theta) >> \hat{f}(x_N,\theta)$, in which case the
posterior on $R_f$ is peaked at $0$. This issue highlights the problem
with using a loudest-event statistic with an improper prior on the
background rate $R_b$. No matter how improbable an event with $x=x_N$
is under the background distribution, it can become likely that the
event at $x_N$ is from the background distribution by taking the
background rate to be sufficiently large. Although this predicts many
more events with $x < x_N$, by using only the loudest event we do not
incorporate the information that no such events are seen. This problem
is avoided in the new framework described here, since we use all
events detected above threshold and combined rates, $R_f+R_b$,
significantly greater than the total number of observed events are
strongly disfavoured.

This problem can be avoided in the loudest event framework, by
including an upper limit on the rate, $R_{\rm max}$, in the prior for
$R_b$. The marginalised distribution for the foreground rate then
becomes
\begin{eqnarray}
p_{\rm LE}\left( R_f, \theta | d \right) &\propto& \left(
\frac{\hat{b}(x_N,\theta)}{(1-\hat{B}(x_N,\theta))^{\frac{3}{2}}}
\left[\frac{\sqrt{\pi}}{2}{\rm
    erf}\left(\sqrt{(1-\hat{B}(x_N,\theta))\,R_{\rm max}}\right) -
  \sqrt{(1-\hat{B}(x_N,\theta))\,R_{\rm max}} \, {\rm
    e}^{-(1-\hat{B}(x_N,\theta)) \, R_{\rm max}}\right]
\right. \nonumber \\ && \left.+ R_f
\hat{f}(x_N,\theta)\frac{\sqrt{\pi}\,{\rm
    erf}\left(\sqrt{(1-\hat{B}(x_N,\theta))\,R_{\rm
      max}}\right)}{\sqrt{(1-\hat{B}(x_N,\theta))}}\right)\frac{p(\theta)}{\sqrt{R_f}}
\exp\left(-R_f(1-\hat{F}(x_N,\theta))\right) \label{LEwithRmax},
\end{eqnarray}
where ${\rm erf}(x)$ is the error function, defined in the usual way
${\rm erf}(x) = (2/\sqrt{\pi}) \int_0^x \exp(-u^2){\rm d}u$. If
$(1-\hat{B}(x_N,\theta)) \, R_{\rm max} \ll 1$, Eq.~(\ref{LEwithRmax})
can be approximated by
\begin{equation}
p_{\rm LE}\left( R_f, \theta | d \right) \propto \left( \frac{R_{\rm
    max}}{3} \hat{b}(x_N,\theta) + R_f
\hat{f}(x_N,\theta)\right)\frac{p(\theta)}{\sqrt{R_f}}
\exp\left(-R_f(1-\hat{F}(x_N,\theta))\right)
\end{equation}
and if $\hat{f}(x_N,\theta) \gg R_{\rm max} \hat{b}(x_N.\theta)$ we
find the same result as before $(1- \hat{F}(x_N,\theta))\,R_f \approx
1/2$.

@@

We can compare these results to results obtained using the two
approximations described earlier, the loudest-event statistic and the
foreground-dominated statistic. The marginalised distribution for the
foreground rate using these alternatives are shown in
Figure~\ref{fig:altmeth}. In this case, the loudest event had
$x_N=9.5$\jon{(correct this)}. The loudest-event statistic depends on
a specification of the maximum, $R_{\rm max}$, for the background
rate. We show results for $R_{\rm max}=\infty$, i.e.m, the improper
prior, and $R_{\rm max}=10000$. The results for other reasonable
choices of $R_{\rm max} = 100, 1000, 100000$ etc. gave exactly the
same posterior, since $\hat{b}(x_N,) R_{\rm max} \ll \hat{f}(x_N)$ for
all these choices and we are therefore in the regime where the
posterior is insensitive to $R_{\rm max}$. To apply the
foreground-dominated statistic we must specify a threshold above which
we assume all events are foreground. It is reasonable to do this based
on a specification for the relative probability of an event being
fore/background, $\hat{f}(x)/\hat{b}(x)=p_{\rm thresh}$. Setting
$p_{\rm thresh}=0.99$ gives $x_{\rm min}=4.07$ and there are $N=13$
events exceeding that threshold. Setting $p_{\rm thresh}=0.5$ gives
$x_{\rm min}=3.82$ and there are $N=15$ events exceeding that
threshold \jon{(correct these numbers using Will's data)}. We show
results for both of these choices of $x_{\rm min}$ in
Figure~\ref{fig:altmeth}. We also show similar results for two extreme
cases --- setting $x_{\rm min}=3.5$ so that we assume all $98$
observed events are foreground; and setting $x_{\rm min}=9$ so that
only the loudest event is classified as foreground.

The loudest event statistic with the improper-prior gives, as
expected, a poor approximation to the foreground rate. The peak is
more accurately located when a prior maximum rate is defined, but the
distribution is much wider than using the full analysis described
here. This is to be expected as much of the information is being
thrown away. The foreground-dominated statistic gives a reasonable
approximation to the true foreground rate, and a distribution that is
not much wider than the full analysis, for the non-extreme choices of
$x_{\rm min}$. For $x_{\rm min}=9$, the foreground dominated statistic
has equivalent performance to the loudest event statistic, as we might
expect, while when $x_{\rm min}=3.5$, it performs poorly since we are
approximating the foreground rate by the total foreground plus
background rate. This indicates that, provided the threshold is chosen
appropriately, the foreground dominated statistic can perform quite
well at estimating the rate. The fact that it reproduces the posterior
from the full analysis so well is indicative of the fact that most of
the information about the foreground comes from the loudest
events. Indeed, for $x_{\rm min} \approx 4$, the set of events
consists mostly of foreground sources and that set contains a decent
number of events ($13$), so it is not surprising that this statistic
can estimate the foreground rate well. The choice of $x_{\rm min}$ was
motivated by knowledge of $\hat{b}(x,\theta)$ and $\hat{f}(x,\theta)$,
which we have in this case, but this might not be possible if the
parameters $\theta$ were simultaneously being estimated. The full
analysis naturally incorporates inference about the rate parameters
$\theta$ and the background rate $R_b$ along with the foreground rate
and incorporates maximum information from the data set and should
therefore lead to narrower posteriors in general.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{pfore}
  \caption{\label{fig:analytic-rate-foreground-probs} Foreground
    probability for each event in the synthetic data set of \S
    \ref{sec:analytic-GW-example} marginalized over all other
    parameters (errorbars are one standard deviation).  True
    foreground events are in dark grey, background events in light
    grey.  Even though our method cannot identify the status of most
    events with confidence, it can still correctly estimate the rates
    (Figure \ref{fig:analytic-rate-recovery}).}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{AltMethodComp}
  \caption{\label{fig:altmeth}Posteriors on foreground rate obtained
    using alternative methods --- the loudest event statistic and the
    foreground dominated statistic. Results are shown for two
    different choices of the maximum a priori background rate, $R_{\rm
      max}$, in the case of the loudest event (``LE'') statistic, and
    for several different choices of the threshold, $x_{\rm min}$, in
    the case of the foreground dominated (``FD'') statistic.}
\end{figure}

@@

If we consider the two alternative methods, the loudest event and
foreground dominated statistics, we will recover the same foreground
distributions as in the case with $N_{\rm eff}$ fixed, which are shown
in Figure~\ref{fig:altmeth}. This is because the parameter $N_{\rm
  eff}$ affects only the background distribution, to which the
foreground-dominated statistic is insensitive, and in the loudest
event case, after marginalisation over $N_{\rm eff}$ we find
$\int_0^{N_{\rm max}} \hat{b}(x_N,N_{\rm eff}) {\rm d}N_{\rm eff} \ll
3 N_{\rm max}/R_{\rm max} \hat{f}(x_N, N_{\rm eff})$ and so we are
still in the foreground-dominated regime in which the loudest event
tells us nothing about the background. Similarly, neither of these
alternative methods can inform us about the value of $N_{\rm eff}$, a
property of the background. \jon{Technically, if we allowed really
  large values of $N_{\rm eff}$ we would get into a regime where the
  loudest event could be background and we'd get something here. But,
  we would need $N_{\rm max} \sim 10^{20}$ and so I don't think we
  should consider these alternatives further in this case.}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{sky_location}
  \caption{\label{fig:sky-loc} Contours of the posterior probability
    distribution for the center of the cluster, $\vec{x}_0$, in the
    example from \S~\ref{sec:star-cluster}.  The center $(x,y) =
    \left(x_0, y_0\right)$ is determined to within about 5\%
    of the structural radius of the cluster, $r_0$ (see
    Eq.~\eqref{eq:true-cluster-parameters}). \ilya{[Given the cluster
        size of 0.1, it seems a bit surprising that the center is off
        by $\sim 0.05$...]} \will{That is $0.1*r_0$, not $0.1$!}}
\end{figure}

>>>>

For the usual $k=1$ case \citep{Biswas2009}, the marginalised
posterior (Eq.~\eqref{eq:rate-shape-posterior}) becomes
\begin{equation}
p_{\rm LE}\left( R_f, R_b, \theta | d \right) \propto \left(R_f
\hat{f}(x_N,\theta) + R_b \hat{b}(x_N,\theta)\right)\exp\left[-\left(
  R_f (1-\hat{F}(x_N,\theta)) + R_b (1-\hat{B}(x_N,\theta)) \right)
  \right] \frac{p(\theta)}{\sqrt{R_f R_b}}.
\end{equation}
where $x_N$ denotes the loudness of the loudest event. Marginalising
over $R_b$ we obtain
\begin{equation}
p_{\rm LE}\left( R_f, \theta | d \right) \propto \left(
\frac{\hat{b}(x_N,\theta)}{2(1-\hat{B}(x_N,\theta))} + R_f
\hat{f}(x_N,\theta)\right)\frac{\sqrt{\pi}\,
  p(\theta)}{\sqrt{1-\hat{B}(x_N,\theta)}\,\sqrt{R_f}}
\exp\left(-R_f(1-\hat{F}(x_N,\theta))\right) .
\end{equation}
This posterior has a maximum in $R_f$ at
\begin{equation}
R_f =
\frac{\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta))\tilde{b}(x_N,\theta)
  +
  \sqrt{\left(\hat{f}(x_N,\theta)-(1-\hat{F}(x_N,\theta))\tilde{b}(x_N,\theta)\right)^2
    -
    4\tilde{b}(x_N,\theta)(1-\hat{F}(x_N,\theta))\hat{f}(x_N,\theta)}}{4
  \hat{f}(x_N,\theta)(1-\hat{F}(x_N,\theta))}.
\end{equation}
where $\tilde{b}(x_N,\theta)
=\hat{b}(x_N,\theta)/(1-\hat{B}(x_N,\theta))$. If
$\tilde{b}(x_N,\theta) \ll 1$, we obtain the result $(1-
\hat{F}(x_N,\theta))\,R_f \approx 1/2$. This can be understood as
the statement that the rate of foreground events with ranking
statistic greater than $x_N$, $(1- \hat{F}(x_N,\theta))\,R_f$, is of
order $1$, as expected. However, $\tilde{b}(x_N,\theta) = -{\rm d}
[\ln(1-\hat{B}(x,\theta))]/{\rm d}x$ and $(1-\hat{B}(x,\theta))
\rightarrow 0$ as $x \rightarrow \infty$, so this term will be
divergent and for many reasonable examples, we will find
$\tilde{b}(x_N,\theta) >> \hat{f}(x_N,\theta)$, in which case the
posterior on $R_f$ is peaked at $0$. This issue highlights the problem
with using a loudest-event statistic with an improper prior on the
background rate $R_b$. No matter how improbable an event with $x=x_N$
is under the background distribution, it can become likely that the
event at $x_N$ is from the background distribution by taking the
background rate to be sufficiently large. Although this predicts many
more events with $x < x_N$, by using only the loudest event we do not
incorporate the information that no such events are seen. This problem
is avoided in the new framework described here, since we use all
events detected above threshold and combined rates, $R_f+R_b$,
significantly greater than the total number of observed events are
strongly disfavoured.

This problem can be avoided in the loudest event framework, by
including an upper limit on the rate, $R_{\rm max}$, in the prior for
$R_b$. The marginalised distribution for the foreground rate then
becomes
\begin{eqnarray}
p_{\rm LE}\left( R_f, \theta | d \right) &\propto& \left(
\frac{\hat{b}(x_N,\theta)}{(1-\hat{B}(x_N,\theta))^{\frac{3}{2}}}
\left[\frac{\sqrt{\pi}}{2}{\rm
    erf}\left(\sqrt{(1-\hat{B}(x_N,\theta))\,R_{\rm max}}\right) -
  \sqrt{(1-\hat{B}(x_N,\theta))\,R_{\rm max}} \, {\rm
    e}^{-(1-\hat{B}(x_N,\theta)) \, R_{\rm max}}\right]
\right. \nonumber \\ && \left.+ R_f
\hat{f}(x_N,\theta)\frac{\sqrt{\pi}\,{\rm
    erf}\left(\sqrt{(1-\hat{B}(x_N,\theta))\,R_{\rm
      max}}\right)}{\sqrt{(1-\hat{B}(x_N,\theta))}}\right)\frac{p(\theta)}{\sqrt{R_f}}
\exp\left(-R_f(1-\hat{F}(x_N,\theta))\right) \label{LEwithRmax},
\end{eqnarray}
where ${\rm erf}(x)$ is the error function, defined in the usual way
${\rm erf}(x) = (2/\sqrt{\pi}) \int_0^x \exp(-u^2){\rm d}u$. If
$(1-\hat{B}(x_N,\theta)) \, R_{\rm max} \ll 1$, Eq.~(\ref{LEwithRmax})
can be approximated by
\begin{equation}
p_{\rm LE}\left( R_f, \theta | d \right) \propto \left( \frac{R_{\rm
    max}}{3} \hat{b}(x_N,\theta) + R_f
\hat{f}(x_N,\theta)\right)\frac{p(\theta)}{\sqrt{R_f}}
\exp\left(-R_f(1-\hat{F}(x_N,\theta))\right)
\end{equation}
and if $\hat{f}(x_N,\theta) \gg R_{\rm max} \hat{b}(x_N.\theta)$ we
find the same result as before, $(1- \hat{F}(x_N,\theta))\,R_f \approx
1/2$.

@@

We can compare these results to results obtained using the two
approximations described earlier, the loudest-event statistic and the
foreground-dominated statistic. The marginalised distribution for the
foreground rate using these alternatives are shown in
Figure~\ref{fig:altmeth}. In this case, the loudest event had
$x_N=9.5$\jon{(correct this)}. The loudest-event statistic depends on
a specification of the maximum, $R_{\rm max}$, for the background
rate. We show results for $R_{\rm max}=\infty$, i.e., the improper
prior, and $R_{\rm max}=10000$. The results for other reasonable
choices of $R_{\rm max} = 100, 1000, 100000$ etc. gave exactly the
same posterior, since $\hat{b}(x_N) R_{\rm max} \ll \hat{f}(x_N)$ for
all these choices and we are therefore in the regime where the
posterior is insensitive to $R_{\rm max}$. To apply the
foreground-dominated statistic we must specify a threshold above which
we assume all events are foreground. It is reasonable to do this based
on a specification for the relative probability of an event being
fore/background, $\hat{f}(x)/\hat{b}(x)=p_{\rm thresh}$. Setting
$p_{\rm thresh}=0.99$ gives $x_{\rm min}=4.07$ and there are $N=13$
events exceeding that threshold. Setting $p_{\rm thresh}=0.5$ gives
$x_{\rm min}=3.82$ and there are $N=15$ events exceeding that
threshold \jon{(correct these numbers using Will's data)}. We show
results for both of these choices of $x_{\rm min}$ in
Figure~\ref{fig:altmeth}. We also show similar results for two extreme
cases --- setting $x_{\rm min}=3.5$ so that we assume all $98$
observed events are foreground; and setting $x_{\rm min}=9$ so that
only the loudest event is classified as foreground.

The loudest event statistic with the improper prior gives, as
expected, a poor approximation to the foreground rate. The peak is
more accurately located when a prior maximum rate is defined, but the
distribution is much wider than using the full analysis described
here. This is to be expected as much of the information is being
thrown away. The foreground-dominated statistic gives a reasonable
approximation to the true foreground rate, and a distribution that is
not much wider than the full analysis, for the non-extreme choices of
$x_{\rm min}$. For $x_{\rm min}=9$, the foreground dominated statistic
has equivalent performance to the loudest event statistic, as we might
expect, while when $x_{\rm min}=3.5$, it performs poorly since we are
approximating the foreground rate by the total foreground plus
background rate. This indicates that, provided the threshold is chosen
appropriately, the foreground dominated statistic can perform quite
well at estimating the rate. The fact that it reproduces the posterior
from the full analysis so well is indicative of the fact that most of
the information about the foreground comes from the loudest
events. Indeed, for $x_{\rm min} \approx 4$, the set of events
consists mostly of foreground sources and that set contains a decent
number of events ($13$), so it is not surprising that this statistic
can estimate the foreground rate well. The choice of $x_{\rm min}$ was
motivated by knowledge of $\hat{b}(x,\theta)$ and $\hat{f}(x,\theta)$,
which we have in this case, but this might not be possible if the
parameters $\theta$ were simultaneously being estimated. The full
analysis naturally incorporates inference about the rate parameters
$\theta$ and the background rate $R_b$ along with the foreground rate
and incorporates maximum information from the data set and should
therefore lead to narrower posteriors in general.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{pfore}
  \caption{\label{fig:analytic-rate-foreground-probs} Foreground
    probability for each event in the synthetic data set of \S
    \ref{sec:analytic-GW-example} marginalized over all other
    parameters (error bars are one standard deviation).  True
    foreground events are in dark grey, background events in light
    grey.  Even though our method cannot identify the status of most
    events with confidence, it can still correctly estimate the rates
    (Figure \ref{fig:analytic-rate-recovery}).}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{AltMethodComp}
  \caption{\label{fig:altmeth}Posteriors on foreground rate obtained
    using alternative methods --- the loudest event statistic and the
    foreground dominated statistic. Results are shown for two
    different choices of the maximum a priori background rate, $R_{\rm
      max}$, in the case of the loudest event (``LE'') statistic, and
    for several different choices of the threshold, $x_{\rm min}$, in
    the case of the foreground dominated (``FD'') statistic. \ilya{I can see the figure on the screen, but can't print it...}}
\end{figure}

@@

If we consider the two alternative methods, the loudest event and
foreground dominated statistics, we will recover the same foreground
distributions as in the case with $N_{\rm eff}$ fixed, which are shown
in Figure~\ref{fig:altmeth}. This is because the parameter $N_{\rm
  eff}$ affects only the background distribution, to which the
foreground-dominated statistic is insensitive, and in the loudest
event case, after marginalisation over $N_{\rm eff}$ we find
$\int_0^{N_{\rm max}} \hat{b}(x_N,N_{\rm eff}) {\rm d}N_{\rm eff} \ll
3 N_{\rm max}/R_{\rm max} \hat{f}(x_N, N_{\rm eff})$ and so we are
still in the foreground-dominated regime in which the loudest event
tells us nothing about the background. Similarly, neither of these
alternative methods can inform us about the value of $N_{\rm eff}$, a
property of the background. \jon{Technically, if we allowed really
  large values of $N_{\rm eff}$ we would get into a regime where the
  loudest event could be background and we'd get something here. But,
  we would need $N_{\rm max} \sim 10^{20}$ and so I don't think we
  should consider these alternatives further in this case.} \ilya{On the other hand, this ties in with the comment at the end of the previous subsection about the difficulty of choosing $x_{min}$ for the foreground-dominated statistic when there are other parameters $\theta$ to be estimated simultaneously ($N_{eff}$ in this case), so perhaps it's worthwhile to add some comment about this?}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{sky_location}
  \caption{\label{fig:sky-loc} Contours of the posterior probability
    distribution for the center of the cluster, $\vec{x}_0$, in the
    example from \S~\ref{sec:star-cluster}.  The center $(x,y) =
    \left(x_0, y_0\right)$ is determined to within about 5\%
    of the structural radius of the cluster, $r_0$ (see
    Eq.~\eqref{eq:true-cluster-parameters}).}
\end{figure}
